---
title: "Intro to Data Science Final Project"
author: "Cova, Langbehn, Villa & Barros" 
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    theme: cosmo
editor: visual
---

# Model Construction and Implementation

------------------------------------------------------------------------
Quitly load in libraries and data seen before (doesn't need to be loaded in twice in main reproducible doc)
```{r message=FALSE, warning=FALSE}
## Load in packages 
library(tidymodels)   
library(tidyverse)   
library(janitor)   
library(naniar)       
library(assertr)      
library(corrplot)
library(gridExtra)

## Turn off scientific notation for readable numbers
options(scipen = 999)

## Read in the CSV file
remittances <- read_csv("../data/remittances.csv") |>
  filter(!is.na(remittances_gdp))

## Set seed for reproducibility 
set.seed(20251211)

## Split data: 80% training, 20% testing
remit_split <- initial_split(data = remittances, prop = 0.8)
remit_train <- training(x = remit_split) 
remit_test <- testing(x = remit_split)
```

## STEP 9: Cross-validation set up
```{r}
## To improve the accuracy of our estimated error rates, we set up a 10-fold cross validation with 5 repetitions since we have a relatively small number of observations within the training data
remit_folds <- vfold_cv(data = remit_train, v = 10, repeats = 5)
```

```{r}
## Recipe (baseline model)
recipe_baseline <- 
  recipe(remittances_gdp ~ stock + gdp_per + unemployment + dist_cap + terror + deportations + internet + inflation,
  data = remit_train) |>
step_impute_median(all_numeric_predictors()) |>
step_impute_mode(all_nominal_predictors())|>
step_mutate(
  gdp_lag = lag(gdp_per), 
  unemp_lag = lag(unemployment)) |>
step_mutate(
    gdp_per = log(gdp_per + 1)) |>
step_normalize(all_numeric_predictors())

## Processing the full training data using parameter specification. 
bake(prep(recipe_baseline, training = remit_train), new_data = remit_train)
```

------------------------------------------------------------------------

## STEP 10: Model Comparison and Evaluation

In this section, we compare five different regression models to predict remittances as a percentage of GDP.

------------------------------------------------------------------------

## 10.1 OLS Baseline Model

```{r}
## Linear model
lm_baseline <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

## Create workflow
lm_workflow <- workflow() |>
  add_recipe(recipe_baseline) |>
  add_model(lm_baseline)

## Fit model
lm_results <- lm_workflow |>
  fit_resamples(resamples = remit_folds)

## Collect RMSE
collect_metrics(lm_results)
```

Establishes a baseline performance using ordinary least squares regression with no regularization. This gives us a benchmark to compare other models against.

------------------------------------------------------------------------

## 10.2 Prepare Enhanced Recipe for Regularized Models

```{r}
library(tidymodels)
library(glmnet)
library(dplyr)

#We clean remittances before folding due to missing values
train_data2 <- remit_train %>%
  filter(!is.na(remittances_gdp))   
remit_folds <- vfold_cv(train_data2, v = 10)

#At first, glmnet was throwing errors, so we need to create a recipe that forces
#your predictors into a form that ridge/lasso (glmnet) can handle, with no NA, no Inf / -Inf
#no constant columns, comparable scales across predictors.

recipe_glmnet <- recipe_baseline %>%
  step_mutate(across(all_numeric_predictors(),
                     ~ if_else(is.finite(.x), .x, NA_real_))) %>% 
  step_impute_median(all_numeric_predictors()) %>%               
  step_impute_mode(all_nominal_predictors()) %>%                 
  step_zv(all_predictors()) %>%                                  
  step_normalize(all_numeric_predictors())                       

ctrl <- control_grid(save_pred = TRUE, verbose = TRUE)
grid30 <- grid_regular(penalty(), levels = 30)

metrics1 <- metric_set(rmse)

#This control object makes errors visible and traceable
```

------------------------------------------------------------------------

## 10.3 RIDGE Regression

```{r}
#this defines ridge regression with a tuned penalty.
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_mode("regression") %>%
  set_engine("glmnet")

# build the workflow (recipe + model)
ridge_wf <- workflow() %>%
  add_recipe(recipe_glmnet) %>%
  add_model(ridge_spec)

#We tune the penalty using cross-validation
ridge_res <- tune_grid(
  ridge_wf,
  resamples = remit_folds,
  grid = grid30,
  metrics = metrics1,
  control = ctrl
)

best_ridge <- select_best(ridge_res, metric = "rmse")
final_ridge_wf <- finalize_workflow(ridge_wf, best_ridge)
ridge_fit <- fit(final_ridge_wf, data = train_data2)

# results
best_ridge
final_ridge_wf
ridge_fit 
```

### Comment:

-   penalty value with the lowest RMSE: best_ridge has a penalty \~ 0. The penalty disappears, with the model becomes almost identical to standard linear regression. The cross-validation procedure
-   shows that adding regularization does not improve predictive performance relative to an unpenalized linear model.
-   ridge_fit shows that ridge ≈ OLS Coefficients will be very similar to baseline linear model

### Conclusion:

The ridge regression regularization indicated increasing deviance as the penalty decreased, with cross-validation selecting a penalty effectively equal to zero. This suggests that the unpenalized linear model already provides an optimal fit for the data.

------------------------------------------------------------------------

## 10.4 LASSO Regression

```{r}
# we specify the LASSO model
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_mode("regression") %>%
  set_engine("glmnet")

# we build the workflow - preprocess to prevent leakage
lasso_wf <- workflow() %>%
  add_recipe(recipe_glmnet) %>%
  add_model(lasso_spec)

# Tune lambda penalty  using cross-validation
lasso_res <- tune_grid(
  lasso_wf,
  resamples = remit_folds,
  grid = grid30,
  metrics = metrics1,
  control = ctrl
)

# choose the best lambda (lowest RMSE)
best_lasso <- select_best(lasso_res, metric = "rmse")
# finalize the workflow
final_lasso_wf <- finalize_workflow(lasso_wf, best_lasso)
# fit the final LASSO model on all training data
lasso_fit <- fit(final_lasso_wf, data = train_data2)

best_lasso
final_lasso_wf
lasso_fit

tidy(lasso_fit) %>%
  filter(term != "(Intercept)") %>%
  filter(estimate != 0) %>%
  arrange(desc(abs(estimate)))
```

### Calculating the RMSE

```{r}
# RMSE comparison
bind_rows(
  collect_metrics(ridge_res) |>
    filter(.metric == "rmse") |>
    inner_join(best_ridge, by = "penalty") |>
    mutate(model = "ridge"),
  collect_metrics(lasso_res) |>
    filter(.metric == "rmse") |>
    inner_join(best_lasso, by = "penalty") |>
    mutate(model = "lasso")
) |>
  select(model, penalty, mean, std_err)
```

### Comment

Cross-validation selected a non-zero penalty for the LASSO model, indicating that it improves predictive performance. The LASSO regularization path shows a small set of predictors entering the model as the penalty decreases, highlighting its role as a variable selection method.

### Comment on selected predictors

The LASSO model selected a set of nine predictors. Remittance intensity is negatively associated with GDP per capita and macroeconomic instability, while unemployment, deportations, and internet access exhibit positive relationships, consistent with counter-cyclical and transaction-cost mechanisms.

### Why LASSO is useful

-   The dataset benefits from variable selection (LASSO), not from coefficient stabilization (RIDGE).

------------------------------------------------------------------------

## 10.5 Random Forest

We use this model a regularizaiton model, to reduce variance error by reducing the important of less important predictors. It is a bagging algorithm which considers each split and divides it into only useful predictors - It uses two hyper paramaters `mtry` which considers x predictors in each split (it can be tuned to optimal value of useful predictors. and `min_n` to stop spliting the data

```{r}
library(vip)
# For missingness inside the dependent variable 
remit_train_clean <- remit_train |>
  filter(!is.na(remittances_gdp))

# Smaller CV for run time optimization 
rf_folds <- vfold_cv(remit_train_clean, v = 5)

## Given the high degree of missingness a recipe that accounts for nas will avoid it from breaking down using median imputation. 
recipe_alt <- 
  recipe(remittances_gdp ~ stock + gdp_per + unemployment + dist_cap + terror + deportations + internet + inflation + `Country Name`,
  data = remit_train_clean) |>
update_role(`Country Name`, new_role = "id") |>
step_impute_median(all_numeric_predictors()) |>
step_lag(gdp_per, unemployment, lag = 1) |>
step_mutate(
    gdp_per = log(gdp_per + 1)) |>
step_normalize(all_numeric_predictors()) 

bake(prep(recipe_alt, training = remit_train_clean), new_data = remit_train_clean)

## Creating a Random Forest Model set up for tuning
rf_mod <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger", 
             importance = "impurity", 
             num.threads = 4)

## Creating a workflow. Need to use alternative specific because it accounts for missingness which will lead the model to fail. 
rf_wf <- workflow() |>
  add_recipe(recipe_alt) |>
  add_model(rf_mod)

## Finalize parameters

rf_params <- rf_wf |> 
  extract_parameter_set_dials() |>
  finalize(remit_train_clean)

rf_params


## tuning grid 
rf_grid <- grid_max_entropy(
  rf_params, 
  size = 20 )

## Tuning it within cross validation using our hyperparamters.
rf_tuned <- rf_wf |>
  tune_grid(
    resamples = rf_folds,
    grid = rf_grid,
    control = control_grid(save_pred = TRUE))

## Measuring the RMSE 
rf_tuned |>
  collect_metrics()

rf_tuned |> 
  show_best(metric = "rmse", n = 10)

## selecting the best specification and fit it to the full training data. 
best_rf <-rf_tuned |>
  select_best(metric = "rmse")

final_rf_wf <- rf_wf |> 
  finalize_workflow(best_rf)

final_rf_fit <- final_rf_wf |> 
  fit(data = remit_train_clean)

# Variable importance 
final_rf_fit |> 
  extract_fit_parsnip() |> 
  vip(num_features = 10)
```

------------------------------------------------------------------------

## 10.6 K-Nearest Neighbors (KNN)

KNN predicts remittances by averaging the values of the K most similar country-year observations.

```{r}
## Create lagged variables by country
remit_train_lagged <- remit_train |>
  arrange(`Country Name`, year) |>
  group_by(`Country Name`) |>
  mutate(
    gdp_lag = lag(gdp_per),
    deportations_lag = lag(deportations)
  ) |>
  ungroup()

## Remove missing values
remit_train_clean_knn <- remit_train_lagged |>
  select(remittances_gdp, stock, gdp_per, unemployment, dist_cap, 
         terror, deportations, internet, inflation, gdp_lag, deportations_lag) |>
  drop_na()

## Cross-validation setup
set.seed(20251211)
knn_folds <- vfold_cv(data = remit_train_clean_knn, v = 10, repeats = 5)

## Recipe: log transform and normalize
recipe_knn <- 
  recipe(remittances_gdp ~ ., data = remit_train_clean_knn) |>
  step_mutate(gdp_per = log(gdp_per + 1)) |>
  step_normalize(all_numeric_predictors())

## Model: KNN with tunable K
knn_mod <- 
  nearest_neighbor(neighbors = tune()) |>
  set_engine("kknn") |>
  set_mode("regression")

## Grid: test K from 1 to 99
knn_grid <- grid_regular(neighbors(range = c(1, 99)), levels = 10)

## Workflow
knn_workflow <- 
  workflow() |>
  add_recipe(recipe_knn) |>
  add_model(knn_mod)

## Tune
knn_results <- 
  knn_workflow |>
  tune_grid(
    resamples = knn_folds,
    grid = knn_grid,
    metrics = metric_set(rmse, rsq)
  )

## Results
knn_results |> collect_metrics()
knn_results |> show_best(metric = "rmse")
knn_results |> autoplot()

## Select best
best_k <- select_best(knn_results, metric = "rmse")
final_knn_wf <- knn_workflow |> finalize_workflow(best_k)
```

------------------------------------------------------------------------

## 10.7 Which Model is the best?

```{r}
## Compare all models
tibble(
  Model = c("OLS", "Ridge", "LASSO", "Random Forest", "KNN"),
  Error = c(
    collect_metrics(lm_results) |> filter(.metric == "rmse") |> pull(mean),
    collect_metrics(ridge_res) |> filter(.metric == "rmse") |> inner_join(best_ridge, by = "penalty") |> pull(mean),
    collect_metrics(lasso_res) |> filter(.metric == "rmse") |> inner_join(best_lasso, by = "penalty") |> pull(mean),
    collect_metrics(rf_tuned) |> filter(.metric == "rmse") |> slice_min(mean) |> pull(mean),
    collect_metrics(knn_results) |> filter(.metric == "rmse") |> slice_min(mean) |> pull(mean)
  )
) |>
  arrange(Error)
```

## Dot plot

```{r}
tibble(
  Model = c("OLS", "Ridge", "LASSO", "Random Forest", "KNN"),
  Error = c(
    collect_metrics(lm_results) |> filter(.metric == "rmse") |> pull(mean),
    collect_metrics(ridge_res) |> filter(.metric == "rmse") |> inner_join(best_ridge, by = "penalty") |> pull(mean),
    collect_metrics(lasso_res) |> filter(.metric == "rmse") |> inner_join(best_lasso, by = "penalty") |> pull(mean),
    collect_metrics(rf_tuned) |> filter(.metric == "rmse") |> slice_min(mean) |> pull(mean),
    collect_metrics(knn_results) |> filter(.metric == "rmse") |> slice_min(mean) |> pull(mean)
  )
) |>
  ggplot(aes(x = Error, y = reorder(Model, -Error))) +
  geom_point(size = 4, color = "steelblue") +
  geom_segment(aes(x = 0, xend = Error, y = Model, yend = Model), color = "gray70") +
  labs(title = "Model Performance", 
       subtitle = "Left is better",
       x = "Prediction Error", 
       y = NULL) +
  theme_minimal()
```

## 10.8 Test the Best Model on New Data

The best model is **Random Forest** (lowest error: 3.06). Now we test it on completely new data.

```{r}
## Create new split for Random Forest (uses clean data)
set.seed(20251211)
rf_split <- initial_split(remit_train_clean, prop = 0.8)

## Test Random Forest
test_results <- final_rf_wf |> last_fit(rf_split)

## Show results
test_results |> collect_metrics()
```

On average, predictions are off by 2.53 percentage points. The model explains 86% of the variation in remittances

```{r}
## Visualize: Actual vs Predicted
test_results |>
  collect_predictions() |>
  ggplot(aes(x = remittances_gdp, y = .pred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest: Test Set Performance",
    subtitle = "Points near line = good predictions",
    x = "Actual Remittances (% GDP)",
    y = "Predicted"
  ) +
  theme_minimal()
```

Random Forest performs **much better** than the other models!
